---
title: "ENVS 193 DS Final"
author: "Claire Timm"
date: "2026-03-11"
format: 
  html: 
    toc: true #display table of contents
    toc-location: right
    toc-depth: 3
    smooth-scroll: true
editor: visual
message: false
warning: false 
---

# Set Up

```{r}
#| label: read in packages and data
#| message: false
#| warning: false

#read in packages here
library(tidyverse)
library(here)
library(janitor)
library(dplyr)
library(readxl)
library(gtsummary)
library(flextable)
library(ggeffects)
library(readr)
library(MuMIn)
library(DHARMa)

#read in data here
kelp <- read_csv("../data/annual_kelp.csv")
mopl <- read_csv("../data/MOPL_nest.csv")
my_data <- read_csv("../data/mydata.csv")
```

# Problem 1. Research writing

## a. [READ WORDS AND WHAT THE METRIC OF CENTRAL TENDENCY!!!]{.underline}

In part one the researchers used a linear regression test to determine whether precipitation predicts flooded wetland area. This test can be used because both predictor variable is precipitation, a continuous variable, and the response variable is flooded wetland area (m^2^), which is also a continuous variable.

In the second part, they likely used a Kruskal-Wallis test to evaluate if median flooded wetland area differs across water year classifications. This test is used to compare medians across more than two groups, in this study the groups were wet, above normal, below normal, dry, or critical drought.

## b.

A box plot could be used to visualize the test in part 2. This graph will have water year classification on the x-axis (wet, above normal, below normal, dry, critical drought) and flooded wetland area (m^2^) on the y-axis. There will be a different colored box plot for each water year classification, each plot will have a median line which will show the statistic being compared in a Kruskal-Wallis test. Additionally there will also be individual data points jittered on the graph to show the distribution of individual observations. The purpose of this graph is to visually allow readers to compare the median flooded wetland area across each of the categories for water year type.

## c.

Based on model predictions, we retained the null hypothesis that precipitation doesn’t predict flooded wetland area (linear regression, F(df, df) = F-statistic, R^2^ = effect size, p = 0.11, ⍺ = significance level).

We retained the null hypothesis and found no difference (η^2^ = stat) in median flooded wetland area (m^2^) across water year classification (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, 𝜒^2^(2)= test stat, p = 0.12, ⍺ = significance level).

## d.

An additional variable that could influence the flooded area of a wetland is water management for wetlands. This is a binary categorical variable (managed vs un-managed). If management actions are taken, such as pumping water or using control gates, this will

# Problem 2. Data visualization

## a.

```{r}
#| label: cleaning data
#| message: false
#| warning: false
kelp_clean <- kelp |> #create new object kelp_clean
  clean_names() |>  #clean up column names (all lowercase and underscores not spaces)
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> #selects only data from these 3 sites
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> #removes a specific observation (probably bad data or an error in collection)
  mutate(site_name = case_when(
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> #creates a new column and changes site codes into names
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |> #changes site_name to a factor and sets the order of site levels 
  group_by(site_name, year) |> #groups data by site and year for summary calculations 
  summarize(mean_fronds = mean(fronds)) |> #calculates the mean number of fronds for each site/year group 
  ungroup() #ungroups data so the data frame can be used normally after 

slice_sample(kelp_clean, n = 5) #randomly displays 5 observations (rows) from the dataset 

str(kelp_clean) #shows the structure of the data frame (variables, data types, and size)
```

## b.

The code chunk above includes a line that filters out frond observations from the date `2000-09-22` in transect 6 in quad 40.

In 1-2 sentences, describe why. What is the number for the value of fronds in that “quad” in that “transect”? What does that value represent? Reference the specific place you found information on what that value represents (e.g. in the Summary tab of the data page).

The certain code chunk is filtered out due to the missing datasheet. In transect 6, quad 40,

## c.

```{r}
#| label: recreating visualization
ggplot(data = kelp_clean, #ggplot call with kelp_clean data frame
       aes(x = year, #year to the x-axis
           y = mean_fronds, #mean_fronds to y-axis
           color = site_name)) +  #colors each data by site_name
  geom_point() + #adds points at each observation to graph 
  geom_line() + #connects data points with a line
  scale_color_manual(values = c( #assigns colors to each site 
    "Naples" = "deeppink4", 
    "Isla Vista" = "slateblue4", 
    "Carpinteria" = "darkgreen")) + 
  labs(
    title = "Sites vary in mean kelp fronds per year", #adds graph title
    x = "Year", #makes x-axis title
    y = "Mean kelp fronds per year", #makes y-axis title
    color = "Site" #adds title to legened 
  ) + 
  theme_minimal() + #changes theme of graph
  theme(
    plot.title = element_text(hjust = 0.5) #centers graph title
  )
```

## d.

Isla Vista is the site that shows the most variable mean kelp fronds per year due to .

Carpinteria shows the least variable mean kelp fronds per year by the .

# Problem 3. Data analysis

## a.

The response variable is the nest survival of a Mountain Plover, the 1s mean the nest successfully survived, while the 0s mean the nest failed, either due to predation or abandonment.

## b.

In this problem, you will be examining the potential effects of *visual obstruction* and *distance to prairie dog colony edge* on the probability of a nest site being occupied by a Mountain Plover.

In 2-3 sentences for each predictor, describe:

-   what kind of variable it is (i.e. continuous, discrete, categorical, etc.),

-   how it could possibly influence the probability of Mountain Plover nest site use (i.e. as visual obstruction increases, the probability of Mountain Plover nest site use would \_\_\_\_\_\_\_\_\_\_\_\_\_\_ because…), and

-   where in the paper you found this information (the specific section, subsection, and paragraph)

Visual Obstruction:

-   This predictor variable is a continuous variable that is measured as

Distance to prairie dog colony:

-   This predictor variable is a continuous variable, measured as the distance (unit) from

## c.

Make a table of all the models you will need to run. You will run 4 models: a null model, a saturated model, and two other models with different combinations of predictors.

Stuck on how to create a table? See workshop 8 for an example.

Your table should have 4 columns: (1) model number, (2) visual obstruction, (3) distance to prairie dog colony, and (4) model description.

| Model number | Visual Obstruction |  Distance to Prairie Dog Colony | Model Description |
|-------------|:-----------:|:-----------:|:--------------------:|
| 1 |  |  | No predictors (null model) |
| 2 | X | X | Saturated model |
| 3 | X |  | Model with visual obstruction only |
| 4 |  | X | Model with distance to prairie dog colony only |

## d.

Then, write your code to run all your models. Do not display any output.

```{r}
#| label: cleaning data up
mopl_clean <- mopl |> 
  clean_names() |> 
  mutate(used_bin = recode(id,
                              "used" = 1, 
                              "available" = 0)) |> 
  select(used_bin, edgedist, vo_nest)

#| label: models
model1 <- glm(used_bin ~ 1, data = mopl_clean, family = binomial)

model2 <- glm(used_bin ~ vo_nest, data = mopl_clean, family = binomial)

model3 <- glm(used_bin ~ edgedist, data = mopl_clean, family = binomial)

model4 <- glm(used_bin ~ vo_nest + edgedist, data = mopl_clean, family = binomial)
```

## e.

```{r}
#| label: AIC model selection
AICc(
  model1,
  model2,
  model3,
  model4
) |> 
  arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) was the model predicting that Mountain Plover nest side was used as a function of visual obstruction. (edit!)

## f.

```{r}
#| label: DHARMa
plot(
  simulateResiduals(model2)
)
```

## g.

```{r}
#| label: data visualization
mod_preds <- ggpredict(model2,
                       terms = "vo_nest")
ggplot(data = mopl_clean,
       aes(x = vo_nest,
           y = used_bin)) +
  geom_point(size = 3, 
             alpha = 0.4,
             color = "deeppink4") + 
  geom_ribbon(data = mod_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.3,
              fill = "lightblue") + 
  geom_line(data = mod_preds,
            aes(x = x, 
                y = predicted)) + 
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 1)) + 
  labs(
    x = "Visual Obstruction at Nest Site (cm)",
    y = "Probability of Mountain Plover Nest Site Use",
    title = "Effect of Visual Obstruction on Mountain Plover Nest Site Use"
    ) + 
  theme_classic()
```

## h.

**Figure 1. Effect of visual obstruction on use of Mountain Plover nest site.** This visualization shows the predicted probability of a Mountain Plover nest site in use as a function of visual obstruction at the nest time (cm). The magenta points represent individual observed data, 0 indicating unused sites and 1 indicating used sites. Due to the transparency of the points, areas of darker circles show more observations at that value. The solid line sloping downwards represents the predictions of a logistic regression model with visual obstruction as the only predictor and the light blue shaded band shows the 95% confidence interval. The visualization shows that the probability of Mountain Plover nest site use decreases as visual obstruction increases. Data source: Duchardt, Courtney; Beck, Jeffrey; Augustine, David (2020). Data from: Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of Black-tailed Prairie Dog disturbance \[Dataset\]. Dryad. <https://doi.org/10.5061/dryad.ttdz08kt7>

## i.

Calculate the predicted probability of Mountain Plover nest site occupancy at 0 on the index of visual obstruction.

Show and annotate all code. Display the output.

```{r}
#| label: probability
ggpredict(model2,
          terms = "vo_nest [0]")
```

(add context?)

## j.

Write 3-5 sentences summarizing what you found, making references to the figure you made in part h and the predictions you calculated in part j. Your summary should include your interpretation of:

-   the predicted probability of nest site occupancy at 0 (no visual obstruction)

-   what relationships, if any, exist between visual obstruction and distance to prairie dog colony and probability of occupancy

-   the biology behind the trends you found - what explains the relationship between visual obstruction and probability of Mountain Plover nest site occupancy?

The predicted probability of nest site occupancy with no visual obstruction (at 0) is .

# Problem 4. Affective and exploratory visualization

## a.

-   How are the visualizations different from each other in the way you have represented your data?

    -   In my Homework 2 exploratory visualization, my data is represented in a factual way that is easy to interpret due to the graph title, axis labels, and different colors for each day of the week, so viewers are able to read the histogram and get a summary of my data. Meanwhile in Homework 3 for my affective visualization, this is a much more abstract form of a visualization, so without my artist statement it would be hard to interpret. Also I created it from my real data but viewers wouldn't be able to get specific data values from looking at it.

-   What similarities do you see between all your visualizations?

    -   Between all my visualizations I can see that Tuesdays on average tend to have the lowest sleep duration, while Monday and Wednesday on average have the most. Additionally I noticed a few outliers on my screen time vs sleep duration plot, but there potentially is a trend for higher sleep duration with a lower screen time.

-   What patterns (e.g. differences in means/counts/proportions/medians, trends through time, relationships between variables) do you see in each visualization? Are these different between visualizations? If so, why? If not, why not?

    -   I noticed that in the affective visualization it was difficult to .

-   What kinds of feedback did you get during week 9 in workshop or from the instructors? How did you implement or try those suggestions? If you tried and kept those suggestions, explain how and why; if not, explain why not.

    -   I received feedback on which colors of yarn to use for my crochet visualization. I took the advice of doing white to represent my sleep duration and found a fluffy type of yarn that reminded me of a blanket. Then for screen time I used a green because I thought the color was pretty, it would stand out, and found it in the same texture as the white one. I took the feedback of instead of using the same color but in two different shades to finding two completely different colors of yarn.

## b.

Based on my predictor variable of daily screen time (minutes) and the response variable, sleep duration (minutes), I could conduct a linear regression analysis to determine the strength, direction, and significance between the two variables. I could figure out if higher amounts of screen time are related to shorter sleep duration since linear regression models the relationship between two variables and since both my predictor and response variable are continuous, measured in minutes, this test will work for quantitative data. Overall this analysis will let me assess if increased screen time will predict changes to sleep duration and allow me to quantify this relationship.

## c.

### Check assumptions

Before I fit a model:

1.  Linear relationship between response and predictor variables

    Yes, I would expect a linear relationship between my X variable, screen time (minutes), and my Y variable, sleep duration (minutes). As screen time increases I would expect sleep duration to decrease.

2.  Independent error (no correlation between errors)

    An error from one observation will not influence an error from another observation since both variables are their own observation and it is made daily.

After I fit my model:

1.  Homoscedasticity of residuals (constant variance)
2.  Normally distributed residuals

```{r}
#| label: diagnostic plots
#clean data 
my_data_clean <- my_data |>  
  clean_names()
#fit the regression model 
model <- lm(sleep_duration_minutes ~ screen_time_minutes, data = my_data_clean)
# base R residuals 
par(mfrow = c (2, 2))
plot(model)
summary(model)
```

[Residuals vs Fitted]{.underline}: This plot checks if there is a linear relationship between the predictor variable, screen time, and the response variable, sleep duration. The residuals appear to be scattered somewhat evenly and randomly around the horizontal line without a strong curved pattern, although there is a slight trend in the red line showing small deviations. Overall the linearity assumption is reasonably statisfied.

[QQ Residuals]{.underline}: This plot checks whether the residuals follow a normal distribution. Since most of the data points fall along the reference line, with the exception of a few, this indicates that the resiudals are normally distributed.

[Scale-Location]{.underline}: This plot checks the assumption of homoscedasticity as well but with the square root of the standardized residuals. The points appear to be relatively evenly spread across the fitted values, suggesting that the variance of residuals is mostly constant so the homoscedasticity assumption is met.

[Residuals vs Leverage]{.underline}: This plot identifies any potential outliers that could affect the regression model using Cook's distance. While some observations appear to be higher than others, none appear to have extremely high influence on the model since they fall within the dashed lines, concluding there are no outliers.

### Analysis

To assess the relationship between screen time (minutes), the predictor variable, and sleep duration (minutes), the response variable, I conducted a simple linear regression analysis. The regression results showed that the estimated intercept is 339.54, which is the predicted sleep duration when screen time is 0. The slope coefficient for screen time is 0.348, meaning that for every increase in 1 minutes of screen time, the predicted sleep duration increases by 0.35 minutes on average. Although the p-value for screen time coefficient is 0.163, which is greater than the significance level of 0.05, indicating a relationship that is not statistically significant.

The F-statistic is 3.04

### Effect size

The R^2^ indicates the proportion of variance in the response variable by the predictor variable. For this model, the R^2^ value is 0.0618 which means that 6.18% of the variation in sleep duration is explained by screen time, a small effect size.

## d.

```{r}
#| label: predictions
data_pred <- ggpredict(
  model,
  terms = "screen_time_minutes"
)
data_pred
```

```{r}
#| label: model summary
tbl_regression (model,
                intercept = TRUE,
                label = list(`(Intercept)` = "Intercept",
                `screen_time_minutes` = "Screen Time")) |> 
  as_flex_table()
```

```{r}
#| label: visualization
ggplot(data = my_data_clean,
       aes(x = screen_time_minutes,
           y = sleep_duration_minutes)) + 
  geom_point(size = 3,
             stroke = 1,
             fill = "purple4",
             shape = 23) +
  geom_ribbon(data = data_pred,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) + 
  geom_line(data = data_pred,
            aes(x = x,
                y = predicted),
            linewidth = 1) + 
  labs(x = "Screen time (minutes)",
       y = "Sleep duration (minutes)",
       title = "Screen Time vs Sleep Duartion") + 
  theme_minimal()
```

## e.

**Figure 2. Screen time was positively associated with sleep duration.** The purple diamonds represent individual data points for each day from 01/26/2026 to 03/11/2026. The predictor variable, screen time which is measured in minutes, is recorded on the x-axis while the response variable, sleep duration which is also measured in minutes, is reflected on the y-axis. The gray shaded band represents the 95% confidence interval around the model predictions. As the number of minutes of screen time increased, the predicted sleep duration also increased based on the fitted linear regression model. The black regression line is slightly sloping upwards, indicating that the higher my daily screen time was, the slightly longer the sleep duration, on average. Although there is quite a bit of variation around the model predictions.

## f. 

Sleep duration (minutes) showed a slight positive relationship with screen time (minutes). Although screen time was not a statistically significant predictor of sleep duration, as screen time increased so did sleep duration (linear regression, F(1, 31) = 2.04, R^2^ = 0.062, p = 0.163, $\alpha$ = 0.05). Based on model predictions, for each additional 1 minute increase in screen time, sleep duration increased by 0.35 minutes (β = 0.348 ± 0.244 SE). On average,

Screen time only explained 6.2% of the variation in sleep duration.

include effect size??
