---
title: "ENVS 193 DS Final"
author: "Claire Timm"
date: "2026-03-11"
format: 
  html: 
    toc: true #display table of contents
    toc-location: right
    toc-depth: 3
    smooth-scroll: true
editor: visual
message: false
warning: false 
---

# Set Up

```{r}
#| label: read in packages and data
#| message: false
#| warning: false

#read in packages here
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(ggeffects)
library(readr)

#read in data here
kelp <- read_csv("../data/SBS_Kelp_All_Years_20250225.csv")
```

# Problem 1. Research writing

## a.

In part one the researchers used a linear regression test to determine whether precipitation predicts flooded wetland area. This is because both variables are continuous, the predictor variable is precipitation and the response variable is flooded wetland area (m^2^).

In the second part, my coworker likely used a Kruskal-Wallis test to evaluate if median flooded wetland area differs across water year classifications. This test is used to compare medians across more than two groups, in this study the groups were wet, above normal, below normal, dry, or critical drought.

## b.

A box plot could be used to visualize the test in part 2. This graph will have water year classification on the x-axis (wet, above normal, below normal, dry, critical drought) and flooded wetland area (m^2^) on the y-axis. There will be a different colored box plot for each water year classification, each plot will have a median line which will show the statistic being compared in a Kruskal-Wallis test. Additionally there will also be individual data points jittered on the graph to show the distribution of individual observations. The purpose of this graph is to visually allow readers to compare the median flooded wetland area across each of the categories for water year type.

## c.

We retained the null hypothesis that precipitation doesn’t predict flooded wetland area (linear regression, , p = 0.11, ⍺ = significance level).

We retained the null hypothesis and found no difference (η^2^ = stat) in median flooded wetland area (m^2^) across water year classification (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, 𝜒^2^(2)= stat, p = 0.12, ⍺ = significance level).

## d.

An additional variable that could influence the flooded area of a wetland is water management for wetlands. This is a binary categorical variable (managed vs un-managed). If management actions are taken, such as pumping water or using control gates, this will

# Problem 2. Data visualization

## a.

```{r}
#| label: cleaning data
#| message: false
#| warning: false
kelp_clean <- kelp |> #create new object kelp_clean
  clean_names() |>  #clean up column names (all lowercase and underscores not spaces)
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> #selects only data from these 3 sites
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> #removes a specific observation (probably bad data or an error in collection)
  mutate(site_name = case_when(
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> #creates a new column and changes site codes into names
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |> #changes site_name to a factor and sets the order of site levels 
  group_by(site_name, year) |> #groups data by site and year for summary calculations 
  summarize(mean_fronds = mean(fronds)) |> #calculates the mean number of fronds for each site/year group 
  ungroup() #ungroups data so the data frame can be used normally after 

slice_sample(kelp_clean, n = 5) #randomly displays 5 observations (rows) from the dataset 

str(kelp_clean) #shows the structure of the data frame (variables, data types, and size)
```

## b.

The code chunk above includes a line that filters out frond observations from the date `2000-09-22` in transect 6 in quad 40.

In 1-2 sentences, describe why. What is the number for the value of fronds in that “quad” in that “transect”? What does that value represent? Reference the specific place you found information on what that value represents (e.g. in the Summary tab of the data page).

## c.

```{r}
#| label: recreating visualization
ggplot(data = kelp_clean, #ggplot call with kelp_clean data frame
       aes(x = year, #year to the x-axis
           y = mean_fronds, #mean_fronds to y-axis
           color = site_name)) +  #colors each data by site_name
  geom_point() + #adds points at each observation to graph 
  geom_line() + #connects data points with a line
  scale_color_manual(values = c( #assigns colors to each site 
    "Naples" = "deeppink4", 
    "Isla Vista" = "slateblue4", 
    "Carpinteria" = "darkgreen")) + 
  labs(
    title = "Sites vary in mean kelp fronds per year", #adds graph title
    x = "Year", #makes x-axis title
    y = "Mean kelp fronds per year", #makes y-axis title
    color = "Site" #adds title to legened 
  ) + 
  theme_minimal() #changes theme of graph
```

## d.

Isla Vista is the site that shows the most variable mean kelp fronds per year due to .

Carpinteria shows the least variable mean kelp fronds per year by the .

# Problem 3. Data analysis

## a.

In 1-2 sentences, explain what the 1s and 0s mean in this data set biologically.

## b.

In this problem, you will be examining the potential effects of *visual obstruction* and *distance to prairie dog colony edge* on the probability of a nest site being occupied by a Mountain Plover.

In 2-3 sentences for each predictor, describe:

-   what kind of variable it is (i.e. continuous, discrete, categorical, etc.),

-   how it could possibly influence the probability of Mountain Plover nest site use (i.e. as visual obstruction increases, the probability of Mountain Plover nest site use would \_\_\_\_\_\_\_\_\_\_\_\_\_\_ because…), and

-   where in the paper you found this information (the specific section, subsection, and paragraph)

## c.

Make a table of all the models you will need to run. You will run 4 models: a null model, a saturated model, and two other models with different combinations of predictors.

Stuck on how to create a table? See workshop 8 for an example.

Your table should have 4 columns: (1) model number, (2) visual obstruction, (3) distance to prairie dog colony, and (4) model description.

## d.

Create a clean dataset to use called `mopl_clean`. In no particular order:

-   clean the column names

-   create a new column called `used_bin` to create 1s and 0s from the `id` column

-   select `used_bin`, `edgedist`, and `vo_nest`

edited 2026-03-11: included instructions for cleaning data

Then, write your code to run all your models. Do not display any output.

## e.

Create a clean dataset to use called `mopl_clean`. In no particular order:

-   clean the column names

-   create a new column called `used_bin` to create 1s and 0s from the `id` column

-   select `used_bin`, `edgedist`, and `vo_nest`

edited 2026-03-11: included instructions for cleaning data

Then, write your code to run all your models. Do not display any output.

## f.

Check your diagnostics for the model you selected using simulated residuals from the `DHARMa` package.

Display all code and output.

## g.

Create a plot showing model predictions *for your selected model* with 95% confidence intervals around the predictions and the underlying data.

Show and annotate all code. Show the output.

For full credit:

-   make sure the x- and y-axis labels are written in full with units if necessary

-   take out the gridlines

-   use colors that are different from the default

## h.

Include a figure number, title, description of the figure, and data citation.

## i.

Calculate the predicted probability of Mountain Plover nest site occupancy at 0 on the index of visual obstruction.

Show and annotate all code. Display the output.

## j.

Write 3-5 sentences summarizing what you found, making references to the figure you made in part h and the predictions you calculated in part j. Your summary should include your interpretation of:

-   the predicted probability of nest site occupancy at 0 (no visual obstruction)

-   what relationships, if any, exist between visual obstruction and distance to prairie dog colony and probability of occupancy

-   the biology behind the trends you found - what explains the relationship between visual obstruction and probability of Mountain Plover nest site occupancy?

# Problem 4. Affective and exploratory visualization

## a.

## b.

## c.

## d.

## e.

## f.

## g.
