---
title: "ENVS 193 DS Final"
author: "Claire Timm"
date: "2026-03-11"
format: 
  html: 
    toc: true #display table of contents
    toc-location: right
    toc-depth: 3
    smooth-scroll: true
editor: visual
message: false
warning: false 
---

# Set Up

```{r}
#| label: read in packages and data
#| message: false
#| warning: false

#read in packages here
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(ggeffects)
library(readr)

#read in data here
kelp <- read_csv("data/SBS_Kelp_All_Years_20250225.csv")
```

# Problem 1. Research writing

## a.

In part one the researchers used a linear regression test to determine whether precipitation predicts flooded wetland area. This is because both variables are continuous, the predictor variable is precipitation and the response variable is flooded wetland area (m^2^).

In the second part, my coworker likely used a Kruskal-Wallis test to evaluate if median flooded wetland area differs across water year classifications. This test is used to compare medians across more than two groups, in this study the groups were wet, above normal, below normal, dry, or critical drought.

## b.

A box plot could be used to visualize the test in part 2. This graph will have water year classification on the x-axis (wet, above normal, below normal, dry, critical drought) and flooded wetland area (m^2^) on the y-axis. There will be a different colored box plot for each water year classification, each plot will have a median line which will show the statistic being compared in a Kruskal-Wallis test. Additionally there will also be individual data points jittered on the graph to show the distribution of individual observations. The purpose of this graph is to visually allow readers to compare the median flooded wetland area across each of the categories for water year type.

## c.

We retained the null hypothesis that precipitation doesn’t predict flooded wetland area (linear regression, , p = 0.11, ⍺ = significance level).

We retained the null hypothesis and found no difference (η^2^ = stat) in median flooded wetland area (m^2^) across water year classification (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, 𝜒^2^(2)= stat, p = 0.12, ⍺ = significance level).

## d.

An additional variable that could influence the flooded area of a wetland is water management for wetlands. This is a binary categorical variable (managed vs un-managed). If management actions are taken, such as pumping water or using control gates, this will

# Problem 2. Data visualization

## a.

```{r}
#| label: cleaning data
#| message: false
#| warning: false
kelp_clean <- kelp |> #create new object kelp_clean
  clean_names() |>  #clean up column names (all lowercase and underscores not spaces)
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> #selects only data from these 3 sites
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> #removes a specific observation (probably bad data or an error in collection)
  mutate(site_name = case_when(
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> #creates a new column and changes site codes into names
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |> #changes site_name to a factor and sets the order of site levels 
  group_by(site_name, year) |> #groups data by site and year for summary calculations 
  summarize(mean_fronds = mean(fronds)) |> #calculates the mean number of fronds for each site/year group 
  ungroup() #ungroups data so the data frame can be used normally after 

slice_sample(kelp_clean, n = 5) #randomly displays 5 observations (rows) from the dataset 

str(kelp_clean) #shows the structure of the data frame (variables, data types, and size)
```

## b.

## c.

## d.

# Problem 3. Data analysis

## a.

## b.

## c.

## d.

## e.

## f.

## g.

## h.

## i.

## j.

# Problem 4. Affective and exploratory visualization

## a.

## b.

## c.

## d.

## e.

## f.

## g.
